Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-27 wpe.tex
new draft of related.tex
various small changes
make WPE section more readable and consistant

For the purposes of this paper, one can view entropy as a measure of complexity and unpredictability in a time series.  A high-entropy time series is almost completely unpredictable---and conversely a low entropy signal is often quite predictable.  This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Shannon entropy rate is equal to the sum of the positive Lyapunov exponents, $\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the larger the entropy and the more difficult prediction becomes.

Further we can consider the information contained in a single observable, say the \emph{present}, of the system, measured by the entropy. This information can be partitioned in to two components: the information shared with past observations --- the mutual information between the past and present --- and the information in the present which is not contained in the past --- the conditional entropy. The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the aforementioned \emph{entropy rate}. In theory, the more redundancy in a signal the more predictable it is. The difference between prediction methods is the types of redundancy (linear, non-linear, etc) they can detect and make use of.

Utilizing entropy as a measure of temporal complexity is by no means a new idea \cite{Shannon1951, mantegna1994linguistic}. Its effective usage requires categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite \emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is effectively\footnote{Measurements from real devices are of finite precision, rendering them discrete. Unfortunately the space of values taken is so large that for the purposes of entropy calculations it can be taken as a continuum.} real-valued.  To get around this, one must discretize the data --- typically achieved by binning.  Unfortunately, this is rarely a good solution to the problem, as the binning of the values introduces an additional dynamic on top of the intrinsic dynamics whose entropy is desired\cite{bollt2001}. The field of symbolic dynamics studies how to discretize a time series in such a way that the intrinsic behavior is not perverted, but these methods are fragile in the face of noise and require further understanding of the underlying system, which defeats the purpose of measuring the entropy in the first place.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural complexity measure for time series" \cite{bandt2002per}.  Permutation entropy employs a method of discretizing real-valued time series that follows the intrinsic behavior of the system under examination.  Rather than looking at the statistics of sequences of values, as is done when computing the Shannon entropy, permutation entropy looks at the statistics of the \emph{orderings} of sequences of values using ordinal analysis. Ordinal analysis of a time series is the process of mapping successive time-ordered elements of a time series to their value-ordered permutation of the same size.  By way of example, if $(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among other things, it is generally robust to observational noise and requires no knowledge of the underlying mechanisms.

@@

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$. For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$:
  \begin{align*}
    p(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
  \end{align*}
  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $\ell \ge 2$ is defined as
  \begin{align*}
    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p(\pi) \log_2 p(\pi)
  \end{align*}

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.  With this in mind, it is common in the literature to normalize permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With this convention, ``low'' entropy is close to 0 and ``high'' entropy is close to 1. Finally, it should be noted that the permutation entropy has been shown to be identical to the Kolmolgorov-Sinai (KS) entropy for many large classes of systems \cite{amigo2012permutation}. In turn, and under very loose conditions, the KS entropy is equal to the Shannon entropy rate. What this normalization allows us to do, then, is approximate the redundancy of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

@@

The weighted probability of a permutation is then:
\begin{align*}
  p_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted probability enhances permutations involved in ``large'' features and demotes permutations which are small in amplitude relative to the features of the time series. The weighted permutation entropy is then:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p_w(\pi) \log_2 p_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, and will be in all the results of this paper.

@@

@@

@@

%    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
%  \caption{The weighted permutation entropy of one run of
%  SVD. The gray bands are regions where the window overlaps
%  regimes. The window size used is 5,000 $\times$ 100,000 instructions
%  and the word length is $6$. For reference the instructions per cycle
%  of \svd are plotted as a ghost behind this plot. Each color on the
%  ghosted time series corresonds to the different regimes as selected
%  by rapid shifts in WPE. From left to right each change in color
%  represents a change in regime for 6 regimes in
%  total.}\label{fig:wwpe}

>>>>

For the purposes of this paper one can view the Shannon entropy, and in particular its growth rate  with respect to word length (the \emph{Shannon entropy rate}), as a measure of complexity and unpredictability in a time series. A high-entropy rate time series is almost completely unpredictable---and conversely a low entropy rate signal is often quite predictable. This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Kolmogorov-Sanai (KS) entropy is equal to the sum of the positive Lyapunov exponents, $\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$. The faster the divergence, the larger the entropy and the more difficult prediction becomes. Further, the KS entropy is defined as the supremum of the Shannon entropy rates of all partitions. The partition which achieves this supremum is known as a \emph{generating partition}.

From a different point of view, we can consider the information (as measured by the Shannon entropy) contained in a single observable, say the \emph{present}, of the system. This information can be partitioned in to two components: the information shared with past observations, the mutual information between the past and present, and the information in the present which is not contained in the past, the conditional entropy of the present given the past. The first part is known as the \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the aforementioned \emph{Shannon entropy rate}. In theory, the more redundancy in a signal the more predictable it is. The difference between prediction methods is the types of redundancy (linear, non-linear, etc) they can detect and make use of.

Utilizing Shannon entropy rate as a measure of temporal complexity is not a new idea \cite{Shannon1951, mantegna1994linguistic}, but these approaches required categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite \emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is effectively\footnote{Measurements from real devices are always discrete though typically of very high resolution, for example \texttt{double} precision. Though these values are discete, for the purposes of entropy calculations they are effectively a continuum.} real-valued.  To get around this, one must discretize the data --- typically achieved by binning.  Unfortunately, this is rarely a good solution to the problem, as the binning of the values introduces an additional dynamic on top of the intrinsic dynamics whose entropy is desired\cite{bollt2001}. The field of symbolic dynamics studies how to discretize a time series in such a way that the intrinsic behavior is not perverted, but these methods are fragile in the face of noise and furthermore require understanding of the underlying system, which defeats the purpose of measuring the entropy to quantify predictability in the first place.

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a ``natural complexity measure for time series'' \cite{bandt2002per}. Permutation entropy employs a method of discretizing real-valued time series that follows the intrinsic behavior of the system under examination.  Rather than looking at the statistics of sequences of values, as is done when computing the Shannon entropy, permutation entropy looks at the statistics of the \emph{orderings} of sequences of values using ordinal analysis. Ordinal analysis of a time series is the process of mapping successive time-ordered elements of a time series to their value-ordered permutation of the same size.  By way of example, if $(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern}, $\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  This method has many features; among other things, it is generally robust to observational noise and requires no knowledge of the underlying mechanisms.

@@

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$. For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$:
  \begin{align*}
    P(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
  \end{align*}
  Where $P(\pi)$ quantifies the probability of an ordinal and $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $\ell \ge 2$ is defined as
  \begin{align*}
    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P(\pi) \log_2 P(\pi)
  \end{align*}

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.  With this in mind, it is common in the literature to normalize permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With this convention, ``low'' PE is close to 0 and ``high'' PE is close to 1. Finally, it should be noted that the permutation entropy has been shown to be identical to the Kolmolgorov-Sinai entropy for many large classes of systems \cite{amigo2012permutation}, so long as observational noise is sufficiently small. As mentioned before, this is equal to the Shannon entropy rate of a generating partition of the system. What this normalization allows us to do, then, is approximate the redundancy of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

@@

The weighted probability of a permutation is then:
\begin{align*}
  P_w(\pi) = \frac{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell}) \cdot \delta(\phi(x_{t:t+\ell}), \pi) }{\displaystyle \sum_{t \le T - \ell} w(x_{t+1:t+\ell})}
\end{align*}
where $\delta(x, y)$ is 1 if $x = y$ and 0 otherwise. Effectively, this weighted probability enhances permutations involved in ``large'' features and demotes permutations which are small in amplitude relative to the features of the time series. The weighted permutation entropy is then:
\begin{align*}
  H_w(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} P_w(\pi) \log_2 P_w(\pi),
\end{align*}
which can also be normalized by dividing by $\log_2(\ell!)$, and will be in all the results of this paper.

@@

% {\color{red}TODO: Put this image back in.}

@@

%% We decided to skip the figure completely, since we took out
%% all stuff about PE as regime-shift detection.  
%% [[should add that as future work, though.]]

@@

%  \caption{The weighted permutation entropy of one run of SVD. The gray bands are regions where the window overlaps regimes. The window size used is 5,000 $\times$ 100,000 instructions and the word length is $6$. For reference the instructions per cycle of \svd are plotted as a ghost behind this plot. Each color on the ghosted time series corresonds to the different regimes as determined by visual inspection. From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
